Double Standards in Judging Collective Action

### 2024-11-05 Documents/manuscript.qmd
2nd revision

---
title: "Double Standards in Judging Collective Action"
header: "DOUBLE STANDARDS"
abstract: |
    Collective action is a powerful force driving social change but often sparks contention about what actions are acceptable means to effect social change. Five studies (total $N = 2,979$) investigated double standards in judging collective action---that is, whether observers judge the same protest actions to be more acceptable depending on who the protesters are and what they are protesting. Two studies used item response theory to develop an instrument of 25 controversial protest actions to measure where people draw the line between acceptable and unacceptable forms of collective action. Three preregistered experiments found no consistent evidence for ingroup bias in terms of social class when judging protests for workers' rights (Experiment 1), in terms of race when judging protests for and against defunding the police (Experiment 2), and in terms of gender when judging protests for and against restricting abortion (Experiment 3). Instead, we found that participants who identified as left-wing/liberal (Experiments 1--3) and rejected system-justifying beliefs (Experiments 1--2) considered the same protest actions more acceptable when a cause aligned with their ideological orientation (for workers' rights, for defunding the police, against restricting abortion) than when it did not (against defunding the police, for restricting abortion). Conservative participants considered the same actions somewhat more acceptable when protesters supported, rather than opposed, restricting abortion (Experiment 3) but considered all protest actions, for and against defunding the police, equally unacceptable (Experiment 2). Our findings have theoretical and practical implications for understanding the often-divided response to social movements.
keywords: collective action, double standards, social identity theory, system justification theory, partisanship, Black Lives Matter, abortion
statement: |
  This research shows that what people consider an acceptable *means* of protest depends on what the *end* of a protest is. That is, observers sometimes judge the *same* controversial protest action (e.g., blocking a road) to be more acceptable when the protesters' cause (e.g., defunding the police) aligns with their own ideological position. Conservatives, who support social order and oppose social change, tend to consider all protest actions less acceptable than progressives. Progressives, in turn, tend to consider disruptive but non-violent actions acceptable means to achieve social change and challenge social injustice. This finding helps explain the often-divided response to recent protest movements.
word-count: "11,654 words"
bibliography: references.bib
csl: apa.csl
execute:
  echo: false
format:
  docx:
    reference-doc: word-template.docx
  pdf:
    keep-tex: true
    pdf-engine: xelatex
    template: apa-template.tex
---

@@

Another explanation for ideology-based double standards is that they are rooted in ideological differences in how conservatives and progressives relate to social change and social inequality which, in turn, are motivated by epistemic, existential, and relational needs [@jost_political_2003; @jost_left_2021]. As motivated social cognition, ideological differences shape how we think and feel about political issues, actions, and events. In this view, ideology-based double standards depend on whether the cause of a protest aligns with someone's ideological positions.

@@

Item response theory is a conceptual and statistical framework for understanding how the characteristics of both items and respondents shape responses to a set of items [@demars_irt_2010]. <!-- As such, it allows us to disentangle how acceptable protest actions are from how accepting respondents are of them and, in so doing, to detect double standards in judging collective action. --> When respondents judge whether each of a set of actions is an acceptable means of protest (1 = *yes*, 0 = *no*), a two-parameter logistic item response theory model estimates responses as a function of several latent (unobserved) parameters: 
$$ \Pr ( y_\textit{ij} = 1 ) = \text{logit}^{-1} ( \alpha_i ( \theta_j + \beta_i ) ) $$
where $\Pr ( y_\textit{ij} = 1 )$ is the probability that respondent $j$ considers action $i$ an acceptable means of protest, $\theta_j$ estimates how *accepting* respondent $j$ is of various protest actions, $\beta_i$ estimates how *acceptable* action $i$ is, and $\alpha_i$ estimates how *discriminating* action $i$ is.

Figure \ref{fig:f1} illustrates those relationships for three hypothetical protest actions. $\theta_{(z)}$ is the *z*-standardized tendency for a participant to be more accepting of various actions so that $\beta_i$ determines the probability that the average respondent ($\theta_{(z)} = 0$) considers an action an acceptable means of protest. Action 2 ($\beta_2 = 1.28$) is more acceptable than Actions 1 and 2 ($\{ \beta_2, \beta_3 \} = 0$) as the average respondent is more likely to consider the former ($\Pr = .78$) than the latter ($\Pr = .50$) acceptable means of protest. Action 3 ($\alpha_3 = 3.60$) is more discriminating than Actions 1 and 2 ($\{ \alpha_1, \alpha_2 \} = 1$) as the slope for the relationship between how accepting a respondent is and how likely they are to consider an action acceptable is steeper for the former action. This means that knowing whether someone considers Action 3 acceptable provides more information about them than knowing whether they consider Actions 1 or 2 acceptable.

```{=latex}
\begin{figure}
\centering
\caption{Item response curves for three hypothetical protest actions}
\includegraphics[scale=1]{../Scale Development/figures/figure-1.png}
\caption*{\textit{Note.} $\theta_{(z)}$ follows a standard normal distribution and is shown as both $z$-scores (bottom) and percentiles (top).}
\label{fig:f1}
\end{figure}
```

In this research, item response theory serves two purposes. First, it provides a statistical framework for scale development. In Studies 1 and 2, we select the most discriminating and informative actions to build an instrument for capturing double standards in judging collective action. Second, it provides a formal and statistical definition of double standards in judging collective action as respondents being more accepting of the same protest actions depending on who the protesters are and what they are protesting. In Experiments 1--3, we operationalize double standards as differences in $\theta_{(z)}$ which, as $\theta_{(z)}$ is $z$-standardized, correspond to Cohen's $d$ effect sizes.

@@

\noindent Participants were then asked to name between five and ten actions that fit that definition. Participants were encouraged to think of actions that varied in how acceptable they were in their opinion. We recoded responses into a smaller set of unique actions, which we supplemented with protest actions from the psychological and political science literature [e.g., @sharp_nonviolent_1973]. This process resulted in 72 actions that varied in how acceptable we would expect them to be. For details, see Supplemental Online Material (SOM).

@@

We estimated a graded response model [@burkner_bayesian_2021; @samejima_graded_1997]---an item response theory model for ordinal response variables---for participants' ratings of how often an action would be an acceptable form of collective action.^[Ratings of how often a given action would be an acceptable form of collective action were strongly and negatively correlated with how disruptive, violent, and extreme participants considered the action to be and how negative they felt about it (Table S3).] Based on the results, we selected the 25 most discriminating, informative, and relevant protest actions to form an instrument for measuring double standards in judging collective action in Experiments 1--3. We selected controversial, and thus diagnostic, actions (e.g., disrupting traffic) over actions that almost all respondents considered acceptable (e.g., signing petitions) or unacceptable (e.g., political violence; Figure \ref{fig:f2}). For detailed ratings and results, see SOM.

@@

By varying the participants' and the protesters' group memberships, we tested the preregistered hypothesis that participants would judge the same protest actions as more acceptable if the protesters were ingroup members protesting for the ingroup's rights than if they were outgroup members protesting for the outgroup's rights. By comparing reactions to working-class and middle-class protesters, we further tested how observers judge collective action by lower-status (i.e. working-class) and higher-status (middle-class) group members.^[For Experiments 1 and 2, we had preregistered the additional hypotheses that all participants would  judge  the  same actions  as  more  acceptable when the protesters were from the higher-status group rather than the lower-status group (Experiment 1) and when the goal of the protest was to support rather than challenge the system (Experiment 2). These hypotheses reflected the assumption that the motivation to justify the system was, to some extent, universal. With hindsight, we no longer think that this assumption is tenable and, therefore, omit the hypotheses based on it from the revised manuscript. We still report the same preregistered analyses and their results.] In non-preregistered analyses, we tested whether participants would judge the same protest actions as more or less acceptable depending on their own ideological positions in terms of political orientation, system-justifying beliefs, and social dominance orientation.^[In Experiment 1, we explored a broader range of operationalizations of political ideology, including social dominance orientation, than in the other experiments.] In this way, Experiment 1 provided a first test of the hypothesized class- and ideology-based double standards in judging collective action.

@@

We preregistered the sample size as well as all hypotheses, inclusion/exclusion criteria, measures, and manipulations ([https://osf.io/24wrx/?view_only=8560eb60286149498f8da48bc9f9d99b](https://osf.io/24wrx/?view_only=8560eb60286149498f8da48bc9f9d99b)). We made all materials, data, and analysis scripts available online ([https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62](https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62)). We followed sample size recommendations for item response theory models [@demars_irt_2010, p. 36], planning to recruit 500 participants.

@@

In the screening survey, participants read an accessible definition of what working-class and middle-class/professional jobs are (for details, see SOM). Participants then answered, among other questions, whether they considered their current job---or the jobs they had had in the past or expected to have in the future---to be a working-class job or a middle-class/professional job. As preregistered, we excluded participants from the lower-status group who did not respond  "working-class job" and participants from the higher-status group who did not respond "middle-class/professional job".

@@

```{=latex}
\begin{figure*}[!t]
\centering
\caption{Results from the preregistered (\textbf{A}) and non-preregistered (\textbf{B}) analyses for Experiment 1}
\includegraphics[scale=1]{../Experiment 1/figures/figure-3}
\caption*{\textit{Note.} HS = Higher Status, LS = Lower Status. $\theta_{(z)}$ is the \textit{z}-standardized tendency to consider more controversial actions acceptable means of protest. (\textbf{A}) $p$ and $n$ are, respectively, the predicted proportion and number of actions a participant would consider acceptable means of protest in each condition. Bars enclose the 95\% most plausible estimates. (\textbf{B}) Ribbons enclose, from the darkest to the lightest shade, the 50\%, 80\%, 95\%, and 99\% most plausible estimates.}
\label{fig:f3}
\end{figure*}
```

@@

We estimated this model in CmdStan [@cmdstanr_2021; @cmdstan_2021] using Bayesian statistical methods. Bayesian inference involves choosing a likelihood function and prior distributions. A likelihood function links the observed data to the model parameters and states how likely the observed data are given different values of said model parameters. Prior distributions state how plausible different values of said model parameters are before considering the observed data. Bayesian inference applies Bayes' theorem to update prior distributions in light of the observed data to produce posterior distributions. In contrast to $p$-values and confidence intervals, posterior distributions have a straight-forward interpretation as stating how plausible different values of the model parameters are given the observed data.

Our model derived the likelihood of participants' responses from a Bernoulli likelihood function with a logistic regression equation linking the two item parameters ($\alpha_i$, $\beta_i$), the one participant parameter ($\theta_j$), and the three regression coefficients to the observed data. To identify the model, we constrained $\theta_j$ to have a mean of zero and constrained $\alpha_i$ to have a fixed mean and to be non-negative. We used partial pooling to estimate $\alpha_i$, $\beta_i$, and $\theta_j$. Our model assigned weakly informative prior distributions [@gelman_prior_2017] to all model parameters.^[Our model assigned $\text{Half-Cauchy} (0, 3)$ prior distributions to the standard deviations of $\alpha_i$, $\beta_i$, and $\theta_j$ and $\text{Normal} (0, 3)$ prior distributions to all other model parameters.] We report point estimates, based on the median of posterior samples, and uncertainty intervals, based on the quantiles of posterior samples, that enclose the 95% most plausible estimates.

@@

Figure \ref{fig:f3} shows (B) the $z$-standardized propensity to consider more controversial protest actions acceptable as a function of the three ideological orientation variables. We found that participants who reported a more right-wing political orientation ($\beta_{xy} = -0.25, [-0.37, -0.15]$) and who expressed more agreement with system-justifying beliefs ($\beta_{xy} = -0.20, [-0.30, -0.10]$) tended to find fewer collective actions to be acceptable. In contrast, we found that, after controlling for the other two variables, social dominance orientation was not associated with participants' judgments about how acceptable various collective actions are ($\beta_{xy} = -0.05, [-0.15, 0.05]$). Overall, these findings suggest that people who are right-wing and endorse system-justifying beliefs tend to find various collective actions to be less acceptable than people who are left-wing and do not endorse system-justifying beliefs.^[We ran additional analyses to explore how these associations differed across experimental conditions. We found uncertainty intervals for these associations to overlap across conditions, suggesting that we did not have enough data to differentiate these varying effects.]

@@

```{=latex}
\begin{figure*}[!t]
\centering
\caption{Results from the preregistered analyses for Experiment 2}
\includegraphics[scale=1]{../Experiment 2/figures/figure-4}
\caption*{\textit{Note.} Against = Protesters oppose defunding the police. For = Protesters support defunding the police. $\theta_{(z)}$ is the \textit{z}-standardized tendency to consider more controversial actions acceptable means of protest. Ribbons enclose, from the darkest to the lightest shade, the 50\%, 80\%, 95\%, and 99\% most plausible estimates.}
\label{fig:f4}
\end{figure*}
```

```{=latex}
\begin{figure*}[!t]
\centering
\caption{Predictions from the preregistered (system justification) and non-preregistered (political orientation) analyses for Experiment 2}
\includegraphics[scale=1]{../Experiment 2/figures/figure-5}
\caption*{\textit{Note.} Against = Protesters oppose defunding the police. For = Protesters support defunding the police. $\theta_{(z)}$ is the \textit{z}-standardized tendency to consider more controversial actions acceptable means of protest.}
\label{fig:f5}
\end{figure*}
```

@@

To test our hypotheses, we estimated a series of two-parameter logistic item response theory models with participants' responses to the question whether they thought each action was an acceptable means of protest as the outcome variable. Our models differed from the models used in Experiment 1 in two ways. First, we estimated the item parameters ($\alpha_i$, $\beta_i$) as *correlated* varying effects. Second, we used varying, rather than fixed, effects to estimate differences between the eight conditions. This resulted in partial pooling---condition-wise estimates were shrunk towards each other---and allowed us to compare conditions without multiple comparison problems [@gelman_why_2012]. Our models assigned weakly informative prior distributions to all model parameters.^[Our model assigned $\text{LKJ} (2)$ prior distributions to the Cholesky-transformed correlation matrices for varying effects and $(\text{Half-})\text{Cauchy} (0, 5)$ prior distributions to all other model parameters.] Figure \ref{fig:f4} shows results from three preregistered models we estimated to test our hypotheses.

@@

Model 3 extended Model 1 by estimating participants' responses as a function of their self-reported support for the cause of the protest. We recoded participants' responses to create a predictor variable that encoded support for defunding the police when protesters supported defunding the police and opposition to defunding the police when protesters opposed defunding the police. As preregistered, we modeled this relationship as a monotonic effect [@burkner_modelling_2020] that estimated the average change in the outcome variable across predictor categories as well as how much of this change occurred between each of the four pairs of adjacent predictor categories.^[Our model assigned a Dirichlet prior, $\alpha = {1, 1, 1, 1}$, to the proportions of the overall change that was expected to occur between each of the four pairs of predictor categories.] Participants who supported the protesters' cause did not, on average, consider the same protest actions to be more acceptable than participants who opposed the protesters' cause (`r with(m3_h3, b_text)`). Ergo, our results did not support the alternative hypothesis that ideology-based double standards can be reduced to support for, or opposition to, the cause of a protest.

@@

We included additional measures, not used in the analyses reported in this paper, including measures of moral conviction [@ryan_reconsidering_2014], gender-related system-justifying beliefs [@jost_exposure_2005], and the updated moral foundations questionnaire [@atari_morality_2022, for details, see SOM]. Within the latter questionnaire, we embedded three further attention checks (e.g., "To show that you are paying attention and giving your best effort, please select 'moderately describes me'.").

@@

```{=latex}
\begin{figure*}[!t]
\centering
\caption{Results from the preregistered analyses for Experiment 3}
\includegraphics[scale=1]{../Experiment 3/figures/figure-6}
\caption*{\textit{Note.} Results show the estimated \textit{z}-standardized tendency ($\theta_{(z)}$) to consider more controversial actions acceptable means to protest for or against restricting abortion as a function of the participants' ideology and gender.}
\label{fig:f6}
\end{figure*}
```

@@

Contradicting our hypothesis of group-based double standards, participants in three preregistered experiments did not show consistent ingroup bias in terms of social class when judging protests for workers' rights (Experiment 1), in terms of race when judging protests for and against defunding the police (Experiment 2), and in terms of gender when judging protests for and against restricting abortion (Experiment 3). Instead, we found that members of advantaged groups---middle-class (<!--$d = 0.33$, -->Experiment 1), White (<!--$d = 0.09$, -->Experiment 2), and male (<!--$d = 0.16$,-->Experiment 3) participants---tended to consider all protest actions more acceptable than members of relatively disadvantaged groups. That is, our findings showed that participants' judgements of protest actions depended on their group memberships---but not in the directions implied by the hypothesized group-based double standards.

@@

In the remainder of this General Discussion, we discuss limiting conditions on our findings as well as theoretical and practical implications for understanding the often divided response to collective action.

## Limitations

Despite its strengths, our research has several limitations that constrain the generalizability of our findings. First, our research examined the hypothesized relationships for three political causes---protecting workers' rights, defunding the police, restricting abortion---and thus provided only limited evidence that our findings generalize beyond those causes. Relying on few stimuli to establish an effect threatens the validity and replicability of research findings and constrains the generalizability of psychological research [@yarkoni_generalizability_2022]. Future research should sample a wider range of causes to address this pervasive but often ignored problem [@judd_treating_2012]. This is particularly important when studying collective action as recent theorizing [@jost_missing_2017] and research [@osborne_protesting_2019] highlighted the importance of differentiating between progressive and conservative causes. Second, our research was based on samples from two Western, Educated, Industrialized, Rich, and Democratic [WEIRD, @henrich_weirdest_2010] contexts---the United Kingdom and the United States---which limits the cross-cultural generalizability of our findings.

>>>>

---
title: "Double Standards in Judging Collective Action"
header: "DOUBLE STANDARDS"
abstract: |
    Collective action is a powerful force driving social change but often sparks contention about what actions are acceptable means to effect social change. We investigated double standards in judging collective action---that is, whether observers judge the same protest actions to be more acceptable depending on who the protesters are and what they are protesting. In two studies, we used item response theory to develop an instrument of 25 controversial protest actions to measure where people draw the line between acceptable and unacceptable forms of collective action. In three preregistered experiments ($N = 2776$), we found no consistent evidence for ingroup bias in terms of social class when judging protests for workers' rights (Experiment 1), in terms of race when judging protests for and against defunding the police (Experiment 2), and in terms of gender when judging protests for and against restricting abortion (Experiment 3). Instead, we found that progressive participants (Experiments 1--3) who rejected system-justifying beliefs (Experiments 1--2) considered the same protest actions more acceptable when a cause aligned with their ideological orientation (for workers' rights, for defunding the police, against restricting abortion) than when it did not (against defunding the police, for restricting abortion). Conservative participants considered the same actions somewhat more acceptable when protesters supported, rather than opposed, restricting abortion (Experiment 3) but considered all protest actions, for and against defunding the police, equally unacceptable (Experiment 2). Our findings have theoretical and practical implications for understanding the often-divided response to social movements.
keywords: collective action, double standards, social identity theory, system justification theory, partisanship, Black Lives Matter, abortion
statement: |
  This research shows that what people consider an acceptable *means* of protest depends on what the *end* of a protest is. That is, observers sometimes judge the *same* controversial protest action (e.g., blocking a road) to be more acceptable when the protesters' cause (e.g., defunding the police) aligns with their own ideological position. Conservatives, who support social order and oppose social change, tend to consider all protest actions less acceptable than progressives. Progressives, in turn, tend to consider disruptive but non-violent actions acceptable means to achieve social change and challenge social injustice. These findings help explain the often-divided response to recent protest movements.
word-count: "12,257 words"
bibliography: references.bib
csl: apa.csl
execute:
  echo: false
format:
  # docx:
  #   reference-doc: word-template.docx
  pdf:
    keep-tex: true
    pdf-engine: xelatex
    template: apa-template.tex
---

@@

Another explanation for ideology-based double standards is that they are rooted in ideological differences in how conservatives and progressives relate to social change and social inequality which, in turn, are motivated by basic psychological needs [@jost_political_2003; @jost_left_2021]. As motivated social cognition, ideological differences shape how we think and feel about political issues, actions, and events. In this view, ideology-based double standards depend on whether the cause of a protest aligns with someone's ideological positions.

@@

Item response theory is a conceptual and statistical framework for understanding how the characteristics of both items and respondents shape responses to a set of items [@demars_irt_2010]. Applied to judging protest actions, this framework allows us to disentangle how *acceptable* different protest actions are from how *accepting* different respondents are of them. In line with the long-standing distinction between normative and non-normative collective action [@wright_responding_1990], it recognizes that some protest actions (e.g., signing petitions) are more acceptable than others (e.g., blocking roads). In line with the hypothesized double standards, it recognizes that some people consider more protest actions acceptable than others. By comparing how accepting respondents are of the *same* protest actions in *different* situations, we can use item response theory to detect double standards in judging collective action.

In this research, respondents judge whether each of a set of more and less controversial protest actions is an acceptable means of protest (1 = *yes*, 0 = *no*). We use two-parameter item response theory models to estimate the probability $\Pr ( y_\textit{ij} = 1 )$ that respondent $j$ considers action $i$ an acceptable means of protest as a function of three latent model parameters:
$$ \Pr ( y_\textit{ij} = 1 ) = \text{logit}^{-1} ( \alpha_i ( \theta_j + \beta_i ) ) $$
where $\theta_j$ estimates how *accepting* respondent $j$ is of various actions, $\beta_i$ estimates how *acceptable* action $i$ is, and $\alpha_i$ estimates how *discriminating* action $i$ is.

```{=latex}
\begin{figure}
\centering
\caption{Item response curves for three hypothetical protest actions}
\includegraphics[scale=1]{../Scale Development/figures/figure-1.png}
\caption*{\textit{Note.} $\Pr ( y_\textit{ij} = 1 )$ is the probability that that respondent $j$ considers action $i$ an acceptable means of protest. $\theta_{(z)}$ is the $z$-standardized tendency for a participant to be more accepting of various actions. It follows a standard normal distribution and is shown as both $z$-scores (bottom) and percentiles (top). $\beta_i$ and $\alpha_i$ capture, respectively, how acceptable and discriminating an action is.}
\label{fig:f1}
\end{figure}
```

Figure \ref{fig:f1} illustrates those relationships for three hypothetical protest actions. $\theta_{(z)}$ is the *z*-standardized tendency for a participant to be more accepting of various actions. It follows a normal distribution so that $\theta_{(z)} = 1.28$ means that a respondent is more accepting than 90\% of respondents. $\beta_i$ determines the probability that the average respondent ($\theta_{(z)} = 0$) considers an action an acceptable means of protest. Action 2 ($\beta_2 = 1.28$) is more acceptable than Actions 1 and 3 ($\{ \beta_1, \beta_3 \} = 0$) as the average respondent is more likely to consider the former ($\Pr = .78$) than the latter ($\Pr = .50$) acceptable means of protest. $\alpha_i$ determines how much the probability that a respondent considers an action acceptable depends on how accepting that respondent is. Action 3 ($\alpha_3 = 3.60$) is more discriminating than Actions 1 and 2 ($\{ \alpha_1, \alpha_2 \} = 1$) as the slope for the relationship between how accepting a respondent is and how likely they are to consider an action acceptable is steeper for the former action. This means that knowing whether someone considers Action 3 acceptable provides more information about them than knowing whether they consider Actions 1 or 2 acceptable.

In this research, item response theory serves two purposes. First, it provides a statistical framework for scale development. In Studies 1 and 2, we select the most discriminating and informative actions to build an instrument for capturing double standards in judging collective action. Second, it provides a formal and statistical definition of double standards in judging collective action as respondents being more accepting of the same protest actions depending on who the protesters are and what they are protesting. In Experiments 1--3, we operationalize double standards as differences in $\theta_{(z)}$ which, as $\theta_{(z)}$ is $z$-standardized, correspond to Cohen's $d$ effect sizes. As $\theta_{(z)}$ determines the probability of respondents judging different protest actions acceptable, it can be converted to the predicted proportion $p$ and number $n$ of actions a respondent considers acceptable. Figure \ref{fig:f3} (A) illustrates these relationships. Figures \ref{fig:f3} (B), \ref{fig:f4}, \ref{fig:f5}, and \ref{fig:f6} show the predicted proportion of actions a participant would consider acceptable as an intuitive and easily interpretable effect size.

@@

\noindent Participants were then asked to name between five and ten actions that fit that definition. Participants were encouraged to think of actions that varied in how acceptable they were in their opinion. We recoded responses into a smaller set of unique actions, which we supplemented with protest actions from the psychological and political science literature [e.g., @sharp_nonviolent_1973]. This process resulted in 72 actions that varied in how acceptable we would expect them to be. For details, see Supplemental Online Material A (SOM-A).

@@

We estimated a graded response model [@burkner_bayesian_2021; @samejima_graded_1997]---an item response theory model for ordinal response variables---for participants' ratings of how often an action would be an acceptable form of collective action.^[Ratings of how often a given action would be an acceptable form of collective action were strongly and negatively correlated with how disruptive, violent, and extreme participants considered the action to be and how negative they felt about it (Table S3).] Based on the results, we selected the 25 most discriminating, informative, and relevant protest actions to form an instrument for measuring double standards in judging collective action in Experiments 1--3. We selected controversial, and thus diagnostic, actions (e.g., disrupting traffic) over actions that almost all respondents considered acceptable (e.g., signing petitions) or unacceptable (e.g., political violence; Figure \ref{fig:f2}). For detailed ratings and results, see SOM-A and Tables S1--S3.

@@

By varying the participants' and the protesters' group memberships, we tested the preregistered hypothesis that participants would judge the same protest actions as more acceptable if the protesters were ingroup members protesting for the ingroup's rights than if they were outgroup members protesting for the outgroup's rights. By comparing reactions to working-class and middle-class protesters, we further tested how observers judge collective action by lower-status (i.e. working-class) and higher-status (middle-class) group members. In non-preregistered analyses, we tested whether participants would judge the same protest actions as more or less acceptable depending on their own ideological positions in terms of political orientation, system-justifying beliefs, and social dominance orientation.^[In Experiment 1, we explored a broader range of operationalizations of political ideology, including social dominance orientation, than in the other experiments.] In this way, Experiment 1 provided a first test of the hypothesized class- and ideology-based double standards in judging collective action.

@@

We preregistered the sample size as well as all hypotheses, inclusion/exclusion criteria, measures, and manipulations ([https://osf.io/24wrx/?view_only=8560eb60286149498f8da48bc9f9d99b](https://osf.io/24wrx/?view_only=8560eb60286149498f8da48bc9f9d99b)). For Experiments 1 and 2, we had preregistered the additional hypothesis that all participants would judge  the  same actions  to be more  acceptable when the protesters were from the higher-status group rather than the lower-status group (Experiment 1) and when the goal of the protest was to support rather than challenge the system (Experiment 2). This hypothesis reflected the assumption that the motivation to justify the system was, to some extent, universal. With hindsight, we no longer think that this assumption is tenable and, therefore, do not discuss the hypothesis based on it. We still follow the same preregistered analysis plan and report all results related to our original hypotheses.

We made all materials, data, and analysis scripts available online ([https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62](https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62)). We followed sample size recommendations for item response theory models [@demars_irt_2010, p. 36], planning to recruit 500 participants.

@@

In the screening survey, participants read an accessible definition of what working-class and middle-class/professional jobs are (for details, see SOM-B). Participants then answered, among other questions, whether they considered their current job---or the jobs they had had in the past or expected to have in the future---to be a working-class job or a middle-class/professional job. As preregistered, we excluded participants from the lower-status group who did not respond  "working-class job" and participants from the higher-status group who did not respond "middle-class/professional job".

@@

```{=latex}
\begin{figure*}[!t]
\centering
\caption{Results from the preregistered (\textbf{A}) and non-preregistered (\textbf{B}) analyses for Experiment 1}
\includegraphics[scale=1]{../Experiment 1/figures/figure-3}
\caption*{\textit{Note.} HS = Higher Status, LS = Lower Status. (\textbf{A}) $\theta_{(z)}$ is the \textit{z}-standardized tendency to consider more controversial actions acceptable means of protest. $\Pr$ and $n$ are, respectively, the predicted proportion and number of actions a participant would consider acceptable means of protest in each condition. Bars enclose the 95\% most plausible estimates. (\textbf{B}) Ribbons enclose, from the darkest to the lightest shade, the 50\%, 80\%, 95\%, and 99\% most plausible estimates.}
\label{fig:f3}
\end{figure*}
```

@@

We estimated this model in CmdStan [@cmdstanr_2021; @cmdstan_2021] using Bayesian statistical methods. Bayesian inference involves choosing a likelihood function and prior distributions. A likelihood function links the observed data to the model parameters and states how likely the observed data are given different values of said model parameters. Prior distributions state how plausible different values of said model parameters are before considering the observed data. Bayesian inference applies Bayes' theorem to update prior distributions in light of the observed data to produce posterior distributions. Unlike $p$-values and confidence intervals, posterior distributions have a straight-forward interpretation as stating how plausible different values of the model parameters are given the observed data. We report point estimates, based on the median of posterior samples, and uncertainty intervals, based on the quantiles of posterior samples, that enclose the 95\% most plausible estimates.

Our model derived the likelihood of participants' responses from a Bernoulli likelihood function with a logistic regression equation linking the two item parameters ($\alpha_i$, $\beta_i$), the one participant parameter ($\theta_j$), and the three regression coefficients to the observed data. To identify the model, we constrained $\theta_j$ to have a mean of zero and constrained $\alpha_i$ to have a fixed mean and to be non-negative. We used partial pooling to estimate $\alpha_i$, $\beta_i$, and $\theta_j$. Our model assigned weakly informative prior distributions [@gelman_prior_2017] to all model parameters: $\text{Half-Cauchy} (0, 3)$ prior distributions to the standard deviations of $\alpha_i$, $\beta_i$, and $\theta_j$ and $\text{Normal} (0, 3)$ prior distributions to all other model parameters.

We report differences between conditions in terms of differences in the estimated tendency for a participant to consider more protest actions acceptable ($\theta_{(z)}$) which, as $\theta_{(z)}$ is $z$-standardized, correspond to Cohen's $d$ effect sizes. Likewise, we report standardized regression coefficients ($\beta_{xy}$) that correspond to the estimated change in the $z$-standardized tendency to consider more protest actions acceptable for each $+1\textit{SD}$ change in the predictor variable. All figures show the predicted proportion of protest actions participants would consider acceptable in each condition based on the estimated models.

@@

Figure \ref{fig:f3} shows (B) how participants' tendency to consider more controversial protest actions acceptable varied as a function of the three ideological orientation variables. We found that participants who reported a more right-wing political orientation ($\beta_{xy} = -0.25, [-0.37, -0.15]$) and who expressed more agreement with system-justifying beliefs ($\beta_{xy} = -0.20, [-0.30, -0.10]$) tended to find fewer collective actions to be acceptable. In contrast, we found that, after controlling for the other two variables, social dominance orientation was not associated with participants' judgments about how acceptable various collective actions are ($\beta_{xy} = -0.05, [-0.15, 0.05]$). Overall, these findings suggest that people who are right-wing and endorse system-justifying beliefs tend to find various collective actions to be less acceptable than people who are left-wing and do not endorse system-justifying beliefs.^[We ran additional analyses to explore how these associations differed across experimental conditions. We found uncertainty intervals for these associations to overlap across conditions, suggesting that we did not have enough data to differentiate these varying effects.]

@@

```{=latex}
\begin{figure*}[!t]
\centering
\caption{Results from the preregistered analyses for Experiment 2}
\includegraphics[scale=1]{../Experiment 2/figures/figure-4}
\caption*{\textit{Note.} Against = Protesters oppose defunding the police. For = Protesters support defunding the police. $\Pr$ is the predicted proportion of actions a participant would consider acceptable means of protest. Ribbons enclose, from the darkest to the lightest shade, the 50\%, 80\%, 95\%, and 99\% most plausible estimates.}
\label{fig:f4}
\end{figure*}
```

```{=latex}
\begin{figure*}[!t]
\centering
\caption{Predictions from the preregistered (system justification) and non-preregistered (political orientation) analyses for Experiment 2}
\includegraphics[scale=1]{../Experiment 2/figures/figure-5}
\caption*{\textit{Note.} Against = Protesters oppose defunding the police. For = Protesters support defunding the police. $\Pr$ is the predicted proportion of actions a participant would consider acceptable means of protest in each condition if they were higher ($+1\text{SD}$) or lower ($-1\text{SD}$) than average in system justification (left) or political conservatism (right).}
\label{fig:f5}
\end{figure*}
```

@@

To test our hypotheses, we estimated a series of two-parameter logistic item response theory models with participants' responses to the question whether they thought each action was an acceptable means of protest as the outcome variable. Our models differed from the models used in Experiment 1 in two ways. First, we estimated the item parameters ($\alpha_i$, $\beta_i$) as *correlated* varying effects. Second, we used varying, rather than fixed, effects to estimate differences between the eight conditions. This resulted in partial pooling---condition-wise estimates were shrunk towards each other---and allowed us to compare conditions without multiple comparison problems [@gelman_why_2012]. Our models assigned weakly informative prior distributions: $\text{LKJ} (2)$ prior distributions to the Cholesky-transformed correlation matrices for varying effects and $(\text{Half-})\text{Cauchy} (0, 5)$ prior distributions to all other model parameters. Figure \ref{fig:f4} shows results from three preregistered models we estimated to test our hypotheses.

@@

Model 3 extended Model 1 by estimating participants' responses as a function of their self-reported support for the cause of the protest. We recoded participants' responses to create a predictor variable that encoded support for defunding the police when protesters supported defunding the police and opposition to defunding the police when protesters opposed defunding the police. As preregistered, we modeled this relationship as a monotonic effect [@burkner_modelling_2020] that estimated the average change in the outcome variable across predictor categories as well as how much of this change occurred between each of the four pairs of adjacent predictor categories. This model assigned a Dirichlet prior, $\alpha = {1, 1, 1, 1}$, to the proportions of the overall change that was expected to occur between each of the four pairs of predictor categories. We found that participants who supported the protesters' cause did not, on average, consider the same protest actions to be more acceptable than participants who opposed the protesters' cause (`r with(m3_h3, b_text)`). Ergo, our results did not support the alternative hypothesis that ideology-based double standards can be reduced to support for, or opposition to, the cause of a protest.

@@

We included additional measures, not used in the analyses reported in this paper, including measures of moral conviction [@ryan_reconsidering_2014], gender-related system-justifying beliefs [@jost_exposure_2005], and the updated moral foundations questionnaire [@atari_morality_2023, for details, see SOM-C]. Within the latter questionnaire, we embedded three further attention checks (e.g., "To show that you are paying attention and giving your best effort, please select 'moderately describes me'.").

@@

```{=latex}
\begin{figure*}[!t]
\centering
\caption{Results from the preregistered analyses for Experiment 3}
\includegraphics[scale=1]{../Experiment 3/figures/figure-6}
\caption*{\textit{Note.} Results show the predicted proportion ($\Pr$) of actions a participant would consider acceptable means of protest for or against restricting abortion as a function of the participants' ideology and gender.}
\label{fig:f6}
\end{figure*}
```

@@

Contradicting our hypothesis of group-based double standards, participants in three preregistered experiments did not show consistent ingroup bias in terms of social class when judging protests for workers' rights (Experiment 1), in terms of race when judging protests for and against defunding the police (Experiment 2), and in terms of gender when judging protests for and against restricting abortion (Experiment 3). Instead, we found that members of advantaged groups---middle-class (<!--$d = 0.33$, -->Experiment 1), White (<!--$d = 0.09$, -->Experiment 2), and male (<!--$d = 0.16$,-->Experiment 3) participants---tended to consider all protest actions more acceptable than members of relatively disadvantaged groups. That is, our findings showed that participants' judgments of protest actions depended on their group memberships---but not in the directions implied by the hypothesized group-based double standards.

@@

In the remainder of this General Discussion, we discuss theoretical and practical implications for understanding the often divided response to collective action as well as limitations that constraint the generality of our findings.
